--- 
title: "Directional Regression Analysis & the DiRA Package"
author: "Ella Foster-Molina"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
description: "Examples of how to perform Directional Regression Analyses using the DiRA package."
---

# Overview

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message =F, warning = F)
```

```{css}
table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}
```


```{r }

# dira_functions path: ~/Dropbox/research/dira/dira_functions_2021-07-19.R
library(htmlwidgets)
library(stargazer)
library(foreign)
library(broom)
library(tidyverse)
library(ggpubr)
library(plotly)

source("~/Dropbox/research/dira/dira_functions_2021-08-10.R")
type <- "html"
```



**Draft version, not for distribution. Please do not share this link.**


This document overviews some ways to use the Directional Regression Analysis package. Consider a multiple linear regression surface defined by 

\begin{align*}
\hat{y} = \hat{\beta}_1 x_1 +  \hat{\beta}_2 x_2 + \hat{\beta}_3 x_3 + ... +  \hat{\beta}_n x_n +  \hat{\beta}_0.
\end{align*}

Many commonly examined nested models within that multiple regression are simply directions in the regression surface. These directions can be usefully analyzed by creating a new axis system that is a rotation of the original axis system. 

These are the core commands for performing these analyses. Additional details about options available for these commands is available at (put link here).

  - stargazer.dira(formula, data, model.names, data.subsets)
      + Takes a formula, dataset, a vector of model names, and any subsets of the data needed. Additional options are available. 
      + Outputs a pretty summary of the rotated and nested models defined in *model.names*. The first two explanatory variables in the formula are used to create the nests and rotations.
  - models.dira(formula, data, model.names, data.subsets)
      + Takes a formula, dataset, a vector of model names, and any subsets of the data needed. Additional options are available.
      + Creates a named list of rotated and nested linear models. You can examine the models within that list in more depth using tools like glance.dira() and tidy.dira(), or extract individual models to evaluate separately.
  - glance.dira(dira_models)
      + Takes a named list of models created by models.dira() 
      + Outputs a tidy overview of each model's statistics
  - tidy.dira(dira_models)
      + Takes a named list of models created by models.dira() 
      + Outputs a tidy overview of the coefficients in each model
  - add_3d.directions(plotly_object, model.names, data.subsets)
      + Takes a plotly object and a vector of model names, as well as additional options.
      + Outputs an interactive 3D plot showing the regression surface as well as all specified nested and rotated models. The command requires x, y, and z variables and a dataset, which can be defined in the main plotly object or within add_3d.directions. The default regression surface is defined by $z \sim x + y$, but interaction term models are also accepted. 
  - add_2d.directions(plotly_object, model.names, data.subsets)
      + Takes a plotly object and a vector of model names, as well as additional options.
      + Outputs a graphic with 2-4 2D subplots showing the nested models and rotated models projected into 2D plots. The command requires x, y, and z variables and a dataset, which can be defined in the main plotly object or within add_2d.directions(). The default regression surface is defined by $z \sim x + y$, but interaction term models are also accepted. Note: The graphics object generated by add_2d.directions cannot be modified by the layout command, although plotly object inputted can include a layout layer. The annotate() command can add annotations to the graphics object. Not all models can be plotted in 2D at the moment.
  - plotlist.2d(plotly_object, model.names, data.subsets)
      + Takes a plotly object and a vector of model names, as well as additional options. The command requires x, y, and z variables and a dataset, which can be defined in the main plotly object or within plotlist.2d(). The default regression surface is defined by $z \sim x + y$, but interaction term models are also accepted. Not all models can be plotted in 2D at the moment.
      + Outputs a list of 2D graphics showing crosssections of the 3D graphic created by add_3d.directions. You can extract individual plots to use as needed.

It is currently possible to examine the following nested and rotated models of the multiple linear regression surface. It is also possible to show directional symmetries between the nested and original models for categorical explanatory variables by applying a nested model the original data subsetted to only include one or two of the categories.

```{r, results ='asis'}

tabl <- " 
| model                 | model name                |  direction of first axis | direction of second axis| line color(s)      |
|-----------------------|:-------------------------:|:------------------------:|:-----------------------:|-------------------:|
| y ~ x1 + x2           | lm_multiple               | x1                       |     x2                  | red & orange       |
| y ~ x1                | lm_y.predby.x1            | x2 predicted by x1       |                         |  light blue        |
| y ~ x1 + x2 residuals | lm_y.predby.x1.w.resids   | x2 predicted by x1       |residuals of x2          | dark blue & blue   |
| y ~ x2                | lm_y.predby.x2            | x1 predicted by x2       |                         | fuchsia            |
| y ~ x2 + x1 residuals | lm_y.predby.x2.w.resids   | x1 predicted by x2       |residuals of x1          | purple  & pink     |
| y ~ PC1               | lm_pc                     | first principal component|                         | light grey         |
| y ~ PC1 + PC2         | lm_pc.w.orth              | first principal component| orthogonal to first axis| black & grey       |
| y ~ new.dir + orth    | lm_new.direction          |   user defined direction | orthogonal to first axis| green & lime green  |
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion

```


